---
title: "eda_fas_1"
author: "Olivia Willard"
format: html
---
## Loading Data
```{r}
#| label: import data 
library(readr)
library(ggplot2)

uds_data <- read_csv(here::here("data/NACC_data/original_data/data_cleaned", "uds.csv")) #questionare data

na.omit(uds_data)

```

## Part 1
1. Travel: did subject have any difficulty or need help with travelling
```{r}
travel_data <- table(uds_data$NACCUDSD, uds_data$TRAVEL)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(TRAVEL))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "TRAVEL") +
  theme_minimal()

addmargins(travel_data)
```
2. Remembering Dates: did subject have any difficulty or need help with remembering appointments, etc.
```{r}
remdates_data <- table(uds_data$NACCUDSD, uds_data$REMDATES)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(REMDATES))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "REMDATES") +
  theme_minimal()

addmargins(remdates_data)
```


3. Paying Attention: did subject have any difficulty or need help with paying attention to or understanding tv programs, etc.
```{r}
payattn_data <- table(uds_data$NACCUDSD, uds_data$PAYATTN)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(PAYATTN))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "PAYATTN") +
  theme_minimal()

addmargins(payattn_data)
```


4. Meal Prepping: did subject have any difficulty or need help with preparing a balanced meal
```{r}
mealprep_data <- table(uds_data$NACCUDSD, uds_data$MEALPREP)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(MEALPREP))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "MEALPREP") +
  theme_minimal()

addmargins(mealprep_data)
```


5. Events: did subject have any difficulty or need help with keeping track of current events?
```{r}
events_data <- table(uds_data$NACCUDSD, uds_data$EVENTS)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(EVENTS))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "EVENTS") +
  theme_minimal()

addmargins(events_data)
```


6. Shopping: did subject have any difficulty or need help with shopping alone for grocerys, clothes, etc.
```{r}
shopping_data <- table(uds_data$NACCUDSD, uds_data$SHOPPING)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(SHOPPING))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "SHOPPING") +
  theme_minimal()

addmargins(shopping_data)
```

7. Games: did subject have any difficulty or need help with playing games or working on hobbys
```{r}
games_data <- table(uds_data$NACCUDSD, uds_data$GAMES)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(GAMES))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "GAMES") +
  theme_minimal()

addmargins(games_data)
```

8. Stove: did subject have any difficulty or need help with heating water, remembering to turn off the stove
```{r}
stove_data <- table(uds_data$NACCUDSD, uds_data$STOVE)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(STOVE))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "STOVE") +
  theme_minimal()

addmargins(stove_data)
```

9. Bills: did subject have any difficulty or need help with paying bills, writing checks, etc.
```{r}
bills_data <- table(uds_data$NACCUDSD, uds_data$BILLS)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(BILLS))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "BILLS") +
  theme_minimal()

addmargins(bills_data)
```

10: Taxes: did subject have any difficulty or need help with assemblin tax records, business affairs, or other papers
```{r}
taxes_data <- table(uds_data$NACCUDSD, uds_data$TAXES)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(TAXES))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "TAXES") +
  theme_minimal()

addmargins(taxes_data)
```



## PART 2

A lot of missing data, like almost over 75% of the data is missing for these variables so probably wont use. However for the chi squared tests that did work it did show there was statistical significance


11: APPSEV: Appetite and eating severity
```{r}
appetite_data <- table(uds_data$NACCUDSD, uds_data$APPSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(APPSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "APPSEV") +
  theme_minimal()

addmargins(appetite_data)
```
12: NITESEV: Nighttime behaviors severity
```{r}
nighttime_data <- table(uds_data$NACCUDSD, uds_data$NITESEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(NITESEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "NITESEV") +
  theme_minimal()

addmargins(nighttime_data)
```
13: MOTSEV: Motor disturbance severity
```{r}
motor_data <- table(uds_data$NACCUDSD, uds_data$MOTSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(MOTSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "MOTSEV") +
  theme_minimal()

addmargins(motor_data)
```
14: IRRSEV: Irritability or lability severity
```{r}
irritability_data <- table(uds_data$NACCUDSD, uds_data$IRRSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(IRRSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "IRRSEV") +
  theme_minimal()

addmargins(irritability_data)
```
15: DISNSEV: Disinhibition severity
```{r}
disinhibition_data <- table(uds_data$NACCUDSD, uds_data$DISNSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(DISNSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "DISNSEV") +
  theme_minimal()

addmargins(disinhibition_data)
```
16: APASEV: apathy or indifference severity
```{r}
apathy_data <- table(uds_data$NACCUDSD, uds_data$APASEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(APASEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "APASEV") +
  theme_minimal()

addmargins(apathy_data)
```
17: ELATSEV: elation or euphoria severity
```{r}
elation_data <- table(uds_data$NACCUDSD, uds_data$ELATSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(ELATSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "ELATSEV") +
  theme_minimal()

addmargins(elation_data)
```
18: ANXSEV: anxiety severity
```{r}
anxiety_data <- table(uds_data$NACCUDSD, uds_data$ANXSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(ANXSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "ANXSEV") +
  theme_minimal()

addmargins(anxiety_data)
```
19: DEPDSEV: depression or dysphoria severity
```{r}
depression_data <- table(uds_data$NACCUDSD, uds_data$DEPDSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(DEPDSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "DEPDSEV") +
  theme_minimal()

addmargins(depression_data)
```
20: AGITSEV: agitation or agression severity
```{r}
agitation_data <- table(uds_data$NACCUDSD, uds_data$AGITSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(AGITSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "AGITSEV") +
  theme_minimal()

addmargins(agitation_data)
```
21: HALLSEV: hallucinations severity
```{r}
hallucination_data <- table(uds_data$NACCUDSD, uds_data$HALLSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(HALLSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "HALLSEV") +
  theme_minimal()

addmargins(hallucination_data)
```
22: DELSEV: delusions severity
```{r}
delusion_data <- table(uds_data$NACCUDSD, uds_data$DELSEV)

ggplot(uds_data, aes(x = NACCUDSD, fill = factor(DELSEV))) +
  geom_bar(position = "fill") +
  labs(x = "NACCUDSD", y = "Proportion", fill = "DELSEV") +
  theme_minimal()

addmargins(delusion_data)
```
