---
title: "Modeling SVM"
author: "Yvette Vargas"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
library(e1071)
library(tidyverse)
library(tidymodels)
library(janitor)

###########################################################################
uds <- read_csv(here::here("data/NACC_data/original_data/data_cleaned/uds.csv"))

udsselect <- c("SEX", "EDUC", "NACCAGE",
        "VEG", "ANIMALS", "TRAILA", "TRAILB", "CRAFTDRE", "MINTTOTS",
        "MEMPROB", "DROPACT", "WRTHLESS", "BETTER", "BORED", "HELPLESS",
        "TAXES", "BILLS", "REMDATES", "TRAVEL", "MEALPREP", "GAMES",
        "NACCUDSD")

udsselect <- c("CRAFTDRE", "TAXES", "REMDATES", "BILLS", "NACCUDSD")

categorical_vars <- c("SEX", "MEMPROB", "DROPACT", "WRTHLESS", "BETTER",
                  "BORED", "HELPLESS", "TAXES", "BILLS", "REMDATES",
                  "TRAVEL", "MEALPREP", "GAMES")

categorical_vars <- c("TAXES", "REMDATES", "BILLS", "NACCUDSD")
```

```{r}
new_uds <- uds %>%
  select(all_of(udsselect)) %>%
  mutate(across(all_of(categorical_vars), as.factor))
```

# HC VS Not HC
```{r}
HC_notHC_set <- new_uds |>
  filter(NACCUDSD == "1" | NACCUDSD == "3" | NACCUDSD == "4") |>
  mutate(cog_status = as.factor(
    case_when(NACCUDSD == 1 ~ "HC",
                              NACCUDSD == 3 ~ "NHC",
                              NACCUDSD == 4 ~ "NHC"))) |>
  select(-NACCUDSD) |>
  na.omit()

# HC_notHC_train |>
#   tabyl(cog_status)
```

When cost is small, margins will be wide, more room for error, more tolerance for misclassification

```{r}
set.seed(777)
svm_split <- initial_split(HC_notHC_set, prop = 0.8)
HC_notHC_train <- training(svm_split)
HC_notHC_test <- testing(svm_split)

################################################
HC_notHC_svm_fit <- svm(cog_status ~ ., data = HC_notHC_train, kernel = "linear", cost = 10, scale = TRUE, cross = 10)

summary(HC_notHC_svm_fit)
# 85.40925 accuracy (with MINTIOTS and CRAFTDRE)
# 85.79004 (udsselect2)
```

Best cost parameter: 1 (with MINTTOTS and CRAFTDRE)
Best cost parameter: 0.01 (without)
Best cost parameter: 100 (udsselect2)
```{r}
tuned <- e1071::tune(svm, cog_status ~ ., data = HC_notHC_train, kernel = "linear", ranges = list(cost=c(0.001, 0.01, 0.1, 1, 10, 100)))

summary(tuned)
```


```{r fitting again with adjusted cost}
# 84.76269 accuracy udsselect2
HC_notHC_svm_fit <- svm(cog_status ~ ., data = HC_notHC_train, kernel = "linear", cost = 100, scale = TRUE, cross = 10)


# 85.59185 accuracy without MINTTOTS and CRAFTDRE
HC_notHC_svm_fit <- svm(cog_status ~ ., data = HC_notHC_train, kernel = "linear", cost = 0.01, scale = TRUE, cross = 10)

HC_notHC_svm_fit
summary(HC_notHC_svm_fit)
```


``` {r}
pred <- predict(HC_notHC_svm_fit, HC_notHC_test, type = "class")
plot(pred)

table(pred, HC_notHC_test$cog_status)

# Convert predictions and actual values to factors
actual <- factor(HC_notHC_test$cog_status, levels = c("HC", "NHC"))

# Calculate the ROC curve
roc_curve <- roc(actual, as.numeric(pred))

# Plot ROC curve
# plot(roc_curve, col = "red")

# Calculate AUC
auc_value <- auc(roc_curve)
print(auc_value)

```

Without MINTOTS and CRAFTDRE:
(cost = 0.01)
85.59185 accuracy
# ROC AUC: 0.8252
pred    HC  NHC
  HC  2544  587
  NHC  177 1476
  
With MINTOTS and CRAFTDRE:
(cost = 1)
85.40925 accuracy
pred HC NHC
HC  836 125
NHC  48 396

udsselect2
(cost = 100)
# ROC AUC: 0.8438
pred   HC NHC
  HC  918 204
  NHC  69 637

# MCI vs. AD
```{r}
MCI_AD_set <- new_uds |>
  filter(NACCUDSD == "3" | NACCUDSD == "4") |>
  mutate(cog_status = as.factor(case_when(NACCUDSD == 3 ~ "MCI",
                              NACCUDSD == 4 ~ "AD"))) |>
  select(-NACCUDSD) |>
  na.omit()

summary(MCI_AD_set)
```

```{r}
set.seed(777)
svm_split2 <- initial_split(MCI_AD_set, prop = 0.8)
MCI_AD_train <- training(svm_split2)
MCI_AD_test <- testing(svm_split2)

################################################
MCI_AD_svm_fit <- svm(cog_status ~ ., data = MCI_AD_train, kernel = "linear", cost = 10, scale = TRUE, cross = 10)

summary(MCI_AD_svm_fit)
# 83.54941 accuracy
# 83.99812 with CRAFTDRE and MINTTOTS
# 81.2193 udsselect2
```

Best cost parameter: 10
Bect cost parameter: 0.01 (udsselect2)
```{r}
tuned2 <- e1071::tune(svm, cog_status ~ ., data = MCI_AD_train, kernel = "linear", ranges = list(cost=c(0.001, 0.01, 0.1, 1, 10, 100)))

summary(tuned2)
```


```{r fitting again with adjusted cost}
MCI_AD_svm_fit <- svm(cog_status ~ ., data = MCI_AD_train, kernel = "linear", cost = 0.01, scale = TRUE, cross = 10)

MCI_AD_svm_fit
summary(MCI_AD_svm_fit)
# 81.2193 accuracy
```


``` {r}
pred <- predict(MCI_AD_svm_fit, MCI_AD_test, type = "class")
plot(pred)
table(pred, MCI_AD_test$cog_status)


# Convert predictions and actual values to factors
actual2 <- factor(MCI_AD_test$cog_status, levels = c("MCI", "AD"))

# Calculate the ROC curve
roc_curve2 <- roc(actual2, as.numeric(pred))

# Plot ROC curve
# plot(roc_curve, col = "red")

# Calculate AUC
auc_value2 <- auc(roc_curve2)
print(auc_value2)
```

Without MINTOTS and CRAFTDRE:
(cost = 0.01)
83.61057 accuracy

pred   AD MCI
  AD  732 159
  MCI 197 957

With MINTTOTS and CRAFTDRE:
# ROC AUC: 0.8391
(cost = 1)
83.9512 accuracy

pred   AD MCI
  AD  139  30
  MCI  43 321

udsselect2
(cost = )
# ROC AUC: 0.8375
pred   AD MCI
  AD  335  68
  MCI  72 391





```{r chatgpt imputing data}
# Assuming you have the data `new_uds`

# Select relevant columns and prepare datasets
HC_notHC_set <- new_uds %>%
  filter(NACCUDSD == "1" | NACCUDSD == "3" | NACCUDSD == "4") %>%
  mutate(cog_status = case_when(
    NACCUDSD == "1" ~ "HC",
    NACCUDSD == "3" ~ "NHC",
    NACCUDSD == "4" ~ "NHC")) %>%
  select(-NACCUDSD)

# Split the dataset
set.seed(777)
svm_split <- initial_split(HC_notHC_set, prop = 0.8)
HC_notHC_train <- training(svm_split)
HC_notHC_test <- testing(svm_split)

# Define and prepare the recipe
recipe <- recipe(cog_status ~ ., data = HC_notHC_train) %>%
  step_impute_median(all_predictors()) %>%  # Impute missing values
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

prepared_recipe <- prep(recipe, training = HC_notHC_train, retain = TRUE)

# Bake (process) the training and test data
HC_notHC_train_processed <- bake(prepared_recipe, new_data = HC_notHC_train)
HC_notHC_test_processed <- bake(prepared_recipe, new_data = HC_notHC_test)

# Fit the SVM model
HC_notHC_svm_fit <- svm(factor(cog_status) ~ ., data = HC_notHC_train_processed, kernel = "linear", cost = 10, scale = FALSE)

# Predict using the test set
pred <- predict(HC_notHC_svm_fit, HC_notHC_test_processed, type = "class")

# Check lengths
length(pred)
length(HC_notHC_test_processed$cog_status)

# Create confusion matrix
confusion_matrix <- table(pred, HC_notHC_test_processed$cog_status)
print(confusion_matrix)

# Convert confusion matrix to data frame for plotting
conf_matrix_df <- as.data.frame(confusion_matrix)

# Plot the results
ggplot(conf_matrix_df, aes(x = pred, y = Freq, fill = HC_notHC_test_processed$cog_status)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Predicted", y = "Frequency", fill = "Actual") +
  theme_minimal() +
  ggtitle("SVM Prediction vs Actual")
```

```{r}

plot(iris)

plot(iris$Sepal.Length, iris$Sepal.Width, col=iris$Species)
plot(iris$Petal.Length, iris$Petal.Width, col=iris$Species)

s<-sample(150, 100)
col<-c("Petal.Length", "Petal.Width", "Species")
iris_train<-iris[s,col]
iris_test<-iris[-s,col]

svmfit <- svm(Species ~., data = iris_train, kernel = "linear", cost = .1, scale = FALSE)
print(svmfit)
# only works with two dimensions and a response
plot(svmfit, iris_train[,col])

 iris_train[,col]
  iris_train
  
tuned <- tune(svm, Species ~., data = iris_train, kernel = "linear", ranges = list(cost=c(0.001,0.01,.1,1,10,100)))
# Will show the optimal cost parameter
summary(tuned)

p <- predict(svmfit, iris_test[,col], type="class")
plot(p)

table(p, iris_test[,3])
mean(p== iris_test[,3])
```
